\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=2.6cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,urlcolor=blue!50!black}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\op}[1]{\ensuremath{\mathsf{#1}}}
\title{A Self-Contained Cubical Type Theory Kernel in Lean~4:\\
Results, Mathematical Significance, and Relevance to\\
Open Problems in Logic and Higher Category Theory}
\author{The FormalizedMathematics Project}
\date{Last updated: 2026-07-11 (living document, updated as the project advances)}
\begin{document}
\maketitle

\begin{abstract}
This project implements, from scratch and with no external mathematical
libraries, a proof assistant kernel for a CCHM-style cubical type theory
inside Lean~4, together with a library of roughly 250 machine-checked
theorems of synthetic homotopy theory, univalent mathematics, and
category theory carried out \emph{inside} the object language.  Highlights
include a univalence axiom that \emph{computes}, the complete quantified
theorem $\pi_1(S^1)\cong\Z$, the identification $\Omega S^1 \equiv \Z$ as
an equality of types, $T^2 \simeq S^1\times S^1$ with every cell
definitional, the general graduate lemma ($\op{isoToEquiv}$), six
higher inductive types including Eilenberg--MacLane spaces $K(G,1)$,
Mac~Lane's pentagon and triangle coherence in $1$-types, and the
2-categorical structure of the category of categories.  This document
records the deliverables, explains their mathematical significance, and
gives an honest assessment of how the platform bears on open problems:
the homotopy hypothesis, the axiomatization of coherence, and the
comparison problem for weak higher categories.
\end{abstract}

\tableofcontents

\section{Overview}

The project consists of two strictly separated layers.

\begin{description}[leftmargin=1.5em]
\item[The kernel (metatheory, in Lean~4).] A bidirectional type checker,
a normalization-by-evaluation (NbE) evaluator, and a de~Morgan interval
algebra implementing the Cohen--Coquand--Huber--M\"ortberg (CCHM) cubical
type theory: dependent products and sums, a cumulative universe hierarchy
$\U_0 : \U_1 : \cdots$, path types $\op{PathP}$, Kan operations
($\op{transp}$ with a constancy rule, $\op{hcomp}$ with face systems),
Glue types, and a growing family of (higher) inductive types.  The
kernel is $\sim$5{,}000 lines of Lean with \emph{no} dependency on
mathlib; every judgment is decided by executable code.
\item[The library (object language).] Roughly 250 definitions and
theorems, each a closed term of the object type theory together with its
type, verified by the kernel at build time.  Nothing in the library is a
Lean theorem \emph{about} the kernel; it is mathematics carried out
\emph{inside} the implemented type theory, in the same sense that the
Cubical Agda library is mathematics inside Cubical Agda.
\end{description}

This two-level architecture --- a strict metatheory (Lean) hosting a
univalent object theory (cubical) --- is itself of foundational interest;
see \S\ref{sec:open}.

\section{The kernel: features and design principles}

\subsection{Core type theory}
\begin{itemize}[leftmargin=1.5em]
\item De~Morgan interval: free distributive lattice with involution on
interval variables; equality decided by canonical antichain
disjunctive-normal forms.
\item NbE with defunctionalized closures: values are first-order data;
$\sim$90 special closure constructors encode the Kan structural rules
(transport in $\Pi$/$\Sigma$/$\op{PathP}$ families, tube transformers for
eliminator/composition commutation, motive codomains of dependent
eliminators).
\item Bidirectional checking; path application $\op{papp}$ carries
endpoint annotations, which is what makes neutral boundary computation
sound under untyped evaluation.
\item Universes with judgmental cumulativity; sort inference.
\item Kan operations: $\op{transp}$ with the \emph{constancy rule}
(transport along a constant family is the identity, decided by an
occurs-check on the semantic family), $\op{hcomp}$ over face systems
given by conjunctions of $(i{=}0/1)$ constraints, derived $\op{hfill}$
and heterogeneous composition.
\item Glue types with $\op{unglue}$; univalence as the \emph{program}
$\op{ua}$; composition-of-universe ($\op{hcomp}$ at $\U$) computed as a
Glue along branch lines glued by a checker-verified $\op{lineEquiv}$; the
general CCHM \S6.2 $\op{transpGlue}$ including fiber corrections.
\end{itemize}

\subsection{Higher inductive types}
Eight (higher) inductive types are implemented, each contributing a new
implementation pattern:
\begin{center}
\begin{tabular}{ll}
$S^1$ & first HIT; lazy endpoint collapse; eliminator/hcomp commutation\\
$\Sigma A$ (suspension) & first \emph{parameterized} HIT; structural transport\\
$T^2$ (torus) & first \emph{2-cell} constructor\\
$\lVert A\rVert$ (prop.\ truncation) & first \emph{recursive} path constructor\\
pushouts & cells carrying function \emph{annotations} (endpoints not recoverable)\\
$A/R$ (set quotients) & \emph{dependent} eliminator; 2-dimensional truncation cell\\
$K(G,1)$ (Eilenberg--MacLane) & composition 2-cell, 3-dimensional truncation cell\\
lists, sums, $\top$, $\bot$, $\N$, $\Z$ & ordinary inductives (in NbE style)
\end{tabular}
\end{center}

\subsection{Design principles discovered}
Three principles were forced on us by bugs and are now systematic:
\begin{enumerate}[leftmargin=1.7em]
\item \textbf{Lazy boundary collapse.}  Eagerly reducing
$\op{loop}\,0\mapsto\op{base}$ (and its relatives) destroys the Kan
structure that $\op{transpGlue}$ later needs; all endpoint collapses
happen in a single \op{force} function, and eliminators match
\emph{unforced} cell values first.  Five independent bugs traced to
violations of this principle.
\item \textbf{Annotation-carrying cells.}  When a cell's boundary
mentions parameters not recoverable from the cell (pushout's
$\op{push}\,c$ has endpoints $\op{inl}(f\,c)$), the constructor value
carries the needed data, and the checker verifies consistency with the
ambient type.
\item \textbf{Off-face junk tolerance.}  Partial elements are computed
unrestrictedly and are meaningless off their faces; every eliminator now
passes such junk through instead of panicking, making the evaluator total
on the values that occur-check walks can reach.
\end{enumerate}

\section{The library: principal results}

\subsection{Path algebra and univalence}
Function extensionality, path induction $J$ (\emph{derived} as a
transport along a connection square, not primitive), the full groupoid
structure of paths (units, associativity, inverses --- the inverse laws by
connection-shaped fillers), two-dimensional path algebra, and the
computation rules one hopes for:
\[
\op{trans}\,\op{refl}\,\op{refl} \longrightarrow \op{refl}
\quad(\text{at concrete types}),\qquad
\op{transport}\,(\op{ua}\,e) \longrightarrow e.\op{fst}.
\]
On top of $\op{ua}$: $\op{uaIdEquiv}:\op{ua}(\op{idEquiv})\equiv\op{refl}$
by a single Glue square, and the $\eta$-rule
$\op{uaEta} : \op{ua}(\op{pathToEquiv}\,p)\equiv p$ by $J$ one universe up
--- one half of the full univalence equivalence, from which
$\op{isSetPathU}$ (paths between sets in $\U$ form a set) follows by a
retract argument.

\subsection{h-levels}
Contractibility, propositions, sets, groupoids; singletons are
contractible; $\op{isPropIsContr}$ (the classical four-tube
$\op{hcomp}$); being an equivalence is a proposition; being a set is a
proposition; closure of h-levels under $\Pi$, $\Sigma$ (with
propositional fibers), retracts; \textbf{Hedberg's theorem} (decidable
equality implies sets) --- whose proof exercises the derived $J$, the
inverse laws, and case analysis with the ``inspect'' idiom;
$\op{setIsoToEquiv}$ (isomorphisms into sets are equivalences, by a
square-filling in the set); and the \textbf{general graduate lemma}
$\op{isoToEquiv}$ with no h-level hypothesis, implementing the classical
$\op{lemIso}$: three $\op{hfill}$s and two correction squares, with
compound faces $(k\wedge j){=}1$ encoded as conjunctions and
$(k\wedge j){=}0$ as split branches.

\subsection{The circle: a complete computing $\pi_1$}
\begin{itemize}[leftmargin=1.5em]
\item The full quantified theorem $\pi_1(S^1)\cong\Z$: both round trips,
$\op{encode}$/$\op{decode}$ with $\op{decode}$ a genuine dependent
function over $S^1$ built from a $\Pi$-level composition (working around
a documented checker incompleteness) and cancellation-path corrections.
\item \emph{Computation}: $\op{winding}(\op{loop}\cdot\op{loop})
\longrightarrow +2$, $\op{winding}(\op{loop}^{-1})\longrightarrow -1$,
etc., running through HCompU and Glue transport --- among the few
implementations in existence where these classical numbers are
\emph{computed} by univalence rather than proved.
\item $\Omega S^1\equiv\Z$ \emph{as an identification of types in the
universe} ($\op{ua}$ of the loop-space equivalence).
\item Negative results: $\op{loop}\not\equiv\op{refl}$, hence $S^1$ is
not a set; $\op{isSet}(\Omega S^1)$ by a retract of $\Z$; $S^1$ is
connected ($\forall x.\,\lVert \op{base}\equiv x\rVert$).
\end{itemize}

\subsection{Integers, arithmetic, quotients}
$\Z$ with the substitution-stable $\op{pos}/\op{negsuc}$ representation
(the strict-cancellation representation was proved
\emph{substitution-unstable} --- a rediscovery, by mechanized failure, of
why Cubical Agda avoids it); commutative group laws all by induction;
decidable equality for $\N$, $\Z$, booleans, letters, lists (with
Hedberg giving second, independent $\op{isSet}$ proofs); the classical
difference construction $(\N\times\N)/{\sim}\;\simeq\;\Z$ exercising the
full quotient workflow (well-definedness, $\op{elimProp}$, equivalence).

\subsection{Category theory in a univalent universe}
Monoids and the monoid $(\Z,+,0)$; universe-indexed precategories (objects
in $\U_{n+1}$, homs in $\U_n$); the category of types (laws definitional
by $\eta$); the fundamental groupoid of a type; functors and natural
transformations; \textbf{functor categories} --- necessarily one universe
level up, with naturality transported as a \emph{proposition} over
hom-sets; vertical and horizontal composition, whiskering, and the
\textbf{interchange law}: the 2-categorical structure of $\mathbf{Cat}$
machine-verified.  Concretely: the delooping $B\Z$ (one object, homs
$\Z$, composition $+$) and the endofunctor category $[B\Z,B\Z]$.

\subsection{Higher inductive theorems}
\begin{itemize}[leftmargin=1.5em]
\item $S^0\simeq\op{Bool}$ and $S^0\equiv\op{Bool}$, with
$\op{transport}$ computing $\op{north}\mapsto\op{true}$; the famous
demonstration $\op{transport}(\op{ua}\,\op{not})\,\op{true}
\longrightarrow \op{false}$.
\item $S^2:=\Sigma S^1$; suspension functorial; the map
$\sigma:S^1\to\Omega S^2$ underlying the generator of $\pi_2(S^2)$.
\item $T^2\cong S^1\times S^1$ (and, via the general graduate lemma and
$\op{ua}$, $T^2\simeq S^1\times S^1$ and $T^2\equiv S^1\times S^1$):
the celebrated cubical proof where the square cell corresponds to the
pair of loops and \emph{every cell of both round trips is definitional}.
\item $\Sigma A\simeq$ pushout of $\top\leftarrow A\rightarrow\top$
(again all-definitional); wedges $A\vee B$ and mapping cones as
one-liners; the figure eight $S^1\vee S^1$ with a rank-2 cover computing
the \emph{abelianized} winding $\pi_1(S^1{\vee}S^1)\to\Z^2$ --- with the
computed fact that this invariant cannot distinguish $LR$ from $RL$,
a machine demonstration that it only sees the abelianization of $F_2$.
\item Propositional truncation: $\op{isProp}\lVert A\rVert$,
functoriality, idempotence $\lVert\lVert A\rVert\rVert\simeq\lVert
A\rVert$; \emph{mere existence} statements ($S^1$ connected, winding
merely surjective); $\lVert A\rVert\simeq A/\!\sim_{\mathrm{total}}$,
identifying truncation as a quotient.
\item Free-group groundwork: reduced words over two generators; the
cancellation prepend $\op{consG}$ proved an equivalence \emph{generically
in the generator}, with case rewriting by \emph{connections}
($e@(i\wedge j)$) instead of $J$; the rank-2 free cover
$\op{helixF_2}$ and the non-abelian winding $\op{windF_2}$ (generator
computations verified; deep composites pending performance work).
\item Eilenberg--MacLane spaces: the $K(G,1)$ HIT with composition
2-cell and truncation 3-cell; $\op{emloop}$ is a group homomorphism (a
$\op{compPath}$-uniqueness cube against the $\op{emcomp}$ cell);
$\op{emloop}\,1\equiv\op{refl}$ from the group laws and the groupoid
laws of paths; the $\op{Codes}$ family over $K(G,1)$ into the
\emph{groupoid of sets} (with $\op{isGroupoidHSet}$ proved via the
carrier-path retraction and $\op{isSetPathU}$); $\op{uaCompMul}$
--- $\op{ua}$ of composed right-multiplications --- verified (108\,s in
the native runner); $\op{encode}$/$\op{decode}$/$\pi_1(K(G,1))\simeq G$
fully constructed, final checks running on the native infrastructure.
\item \textbf{The two circles are one}: $S^1\simeq K(\Z,1)$ fully
constructed --- $\op{toEM}$ ($\op{loop}\mapsto\op{emloop}\,1$),
$\op{fromEM}$ (recursor with the $\op{intLoopComp}$-corrected
composition cell), and both round trips; the key lemma
$\op{toIntLoop}$ ($\op{cong}\,\op{toEM}$ distributes over composition
\emph{definitionally}, because constant-motive eliminators commute with
$\op{hcomp}$) verifies in 3\,ms natively.  The heavy round-trip checks
run on the native infrastructure.
\end{itemize}

\subsection{Coherence, chapter 1}
Mac~Lane's \textbf{pentagon} and \textbf{triangle} for path composition,
stated with the full associator/whiskering composites and proved
\emph{in any 1-type} (parallel 2-cells in a groupoid agree), with the
$\Z$-instance.  A finding of independent interest: the \emph{untruncated}
pentagon cannot be proven by the naive quadruple path induction, because
$\op{trans}\,\op{refl}\,\op{refl}$ reduces to $\op{refl}$ only at
concrete types; at abstract types the composite is a stuck $\op{hcomp}$,
so the coherence cells ($\op{assocRefl}$, unit-cell collapses, \dots)
must be bootstrapped by hand --- a precise computational location of the
difficulty that coherence theory addresses.

\subsection{Coherence, chapters 2--4: untruncated tools and
Eckmann--Hilton}
Beyond the 1-type case, the library now proves coherence cells
\emph{in an arbitrary type}, with connections only (no $J$, no extra
Kan compositions):
\begin{itemize}[leftmargin=1.5em]
\item \textbf{Definitional endpoint fillers.}
  $\op{transFill} : \op{PathP}\,(\lambda k.\,w \equiv q\,k)\,p\,(p
  \cdot q)$ and its left counterpart, engineered so that an
  identically-true tube face makes the endpoints collapse
  \emph{definitionally}; the untruncated associator
  $\op{assocConn} : p\cdot(q\cdot r) \equiv (p\cdot q)\cdot r$
  then assembles by composing the two fillers across the \emph{moving
  middle point} $q\,k$, with no cube reasoning escaping the fillers.
\item \textbf{A definitional unit coincidence.}
  $\op{transReflR}\,\op{refl}$ and $\op{transReflL}\,\op{refl}$ are
  \emph{convertible} (the 3-cell is $\op{refl}$): the two unit squares
  have all-constant tubes at $\op{refl}$.  In book HoTT this coincidence
  is a nontrivial lemma; here the kernel computes it away.
\item \textbf{The square-exchange principle.}  For any syntactic square
  $S\,k\,m$, the two boundary routes agree:
  $(\mathrm{bottom}\cdot\mathrm{right}) \equiv
  (\mathrm{left}\cdot\mathrm{top})$, proved by sliding the middle
  point along the anti-diagonal $S\,(\neg m)\,m$ with connection
  half-squares.  Instances: the $\op{cong}_2$ exchange law
  ($\op{congSlide}$) and the naturality of both unitors.
\item \textbf{Eckmann--Hilton.}  $\alpha \cdot \beta \equiv \beta
  \cdot \alpha$ for 2-loops in \emph{any} type --- the classical first
  theorem of synthetic homotopy theory beyond $\pi_1(S^1)$ --- as an
  18-step composite of 3-cells: exchange the two loops with
  $\op{congSlide}$, restore whiskered forms with the unitor
  naturalities, and cancel the shared unitor (the definitional unit
  coincidence keeps the left/right bookkeeping inside conversion).
  Verified in milliseconds by the type checker.  Packaged
  algebraically: \emph{the second loop space of any type is a
  commutative monoid} (untruncated), with multiplication composition,
  commutativity Eckmann--Hilton — instantiated at every rung of the
  loop-space tower.
\item \textbf{Canonical face literals and Mac~Lane's triangle.}  A
  kernel improvement (canonical atomic face literals in system
  normalization: $\neg k{=}1 \mapsto k{=}0$, conjunctive splitting)
  makes the fillers of $\op{assocConn}\,p\,\op{refl}\,q$
  \emph{convertible} with the two unitors --- the associator at a middle
  $\op{refl}$ \emph{is} the diagonal of the unitor square.  Mac~Lane's
  \textbf{triangle} then holds untruncated, in any type, by a three-step
  argument (a diagonal-vs-edges lemma by the moving-middle method, plus
  right cancellation).
\item \textbf{Mac~Lane's pentagon, untruncated.}  The full pentagon
  \[ \alpha_{p,q,r\cdot s} \cdot \alpha_{p\cdot q,r,s} \;\equiv\;
     (p\cdot\alpha_{q,r,s}) \cdot \alpha_{p,q\cdot r,s} \cdot
     (\alpha_{p,q,r}\cdot s) \]
  holds \emph{in any type}, with $\alpha := \op{assocConn}$.  Proof
  architecture: three naturality squares obtained by \emph{moving the
  associator along its own fillers} (each a one-line application of the
  square-exchange principle), a \emph{fusion cell} built with
  conjunction faces --- a face $(l{=}0)\wedge(m{=}1)$ is identically
  false on one side of the interpolation and normalizes to $(l{=}0)$ on
  the other, so cells whose face systems differ across the two ends
  become constructible --- and a final \emph{box-with-lid} cell in which
  all four edges of the connecting cube are collapsed by identically-true
  faces whose walls are the fillers themselves.  Supporting kernel work:
  the checker's face resolution was generalized to arbitrary interval
  expressions via canonical-DNF disjunct covers.  Together with the
  triangle this completes \textbf{Mac~Lane's coherence pair, untruncated,
  axiom-free}.
\item \textbf{Dimension-1 homotopy hypothesis: the counit.}  For any
  pointed 1-type $(A,a)$ the loop space $\Omega(A,a)$ carries a group
  structure in which \emph{every law is one of the untruncated coherence
  cells above} (associativity is $\op{assocConn}$, the units are the two
  unitors, the inverse laws are $\op{cancelL}$); the 1-type hypothesis
  enters only through set-ness.  The realization map
  $K(\Omega(A,a),1) \to A$ is `em1rec` with the \emph{identity} loop
  map, whose coherence cell is exactly the composition filler
  $\op{transFill}$ --- and on loops it computes to the identity
  \emph{definitionally} ($\op{refl}$ proof).  This generalizes
  $S^1 \simeq K(\Z,1)$ and opens the programme
  $A \simeq K(\pi_1 A,1)$ for pointed connected 1-types.
\item \textbf{The fundamental groupoid, functorially.}  Internal
  (multi-object) groupoid structure is defined; every 1-type carries it
  on its path spaces with \emph{every} law an instance of the
  untruncated coherence cells, the 1-type hypothesis entering only
  through set-ness of the homs.  The theory is developed functorially:
  groupoid morphisms, identity and composition of morphisms, and the
  functorial action of any map ($F_1 = \op{cong}$, identity
  preservation definitional, composition preservation by
  $\op{congTrans}$).  Two strictness theorems hold \emph{by} `refl`:
  the vertex group of the fundamental groupoid \emph{is} the loop group,
  and $\op{cong}(g \circ f) \equiv \op{cong} g \circ \op{cong} f$.
  With groups as one-object groupoids this completes the
  1-type-to-groupoid direction of the ``groupoids $\simeq$ 1-types''
  roadmap item; the reverse direction's prerequisite — a
  \emph{general-groupoid Eilenberg--MacLane HIT} $B\mathcal{G}$ — is
  now implemented in the kernel: formation, point/arrow/composition/
  truncation constructors, and a recursor that computes on points, with
  smoke tests over $B(\mathrm{Unit},\Z,+)$ all kernel-checked and the
  entire extension validated behaviour-preserving by the differential
  harness.  On this foundation the \textbf{dimension-1 classification
  theorem} is now \emph{fully machine-verified, with no pointedness or
  connectedness hypotheses}:
  \[ B(\Pi_1 A) \;\simeq\; A \qquad\text{(hence } B(\Pi_1 A)
     \equiv A\text{ by univalence)} \]
  for \emph{every} 1-type $A$.  The proof is light precisely because the
  HIT carries the right universal property: the counit is the recursor
  with the composition filler as its coherence cell and computes
  definitionally on points and arrows; the unit is the dependent
  eliminator with `barr ≡ cong bpt` (proved by cancellation and $J$) as
  its arrow cell — no encode--decode, no transport along `ua` lines.
\end{itemize}

\section{Engineering results of independent interest}
\begin{itemize}[leftmargin=1.5em]
\item \textbf{Native versus interpreted checking.}  Build-time guards run
in Lean's elaborator interpreter; moving deep checks to a compiled test
runner yields $\sim$100$\times$: the check of $\op{uaCompMul}$ went from
$>6.75$ hours (unfinished) to 108 seconds.  \texttt{implemented\_by}
native overrides are silently inert in the elaborator --- a fact that had
invalidated several earlier optimization experiments and is worth knowing
for any Lean-hosted evaluator.
\item \textbf{Definition environments.}  References to already-verified
definitions are \emph{trusted} (checked once, in dependency order), the
standard architecture of proof assistants, implemented as a one-node
extension; and an alternative environment-injection mode in which
references resolve to de~Bruijn variables over a preloaded, value-shared
context.
\item \textbf{A catalogue of failure modes} for evaluator caching
(GC address reuse causing unsound hits; capture-based over-approximation
in occurs checks breaking \emph{load-bearing} definitional equalities),
each documented with its mechanism.
\end{itemize}

\section{Mathematical significance}

\begin{enumerate}[leftmargin=1.7em]
\item \textbf{Univalence computes.}  The kernel is a working
demonstration --- independent of Cubical Agda and \texttt{cubicaltt} ---
that the univalence axiom can be a program: transport along
$\op{ua}(e)$ \emph{evaluates} to $e$, winding numbers are computed, and
the classical $\op{transport}(\op{ua}\,\op{not})$ example runs.  This is
evidence in the tradition of constructive canonicity for cubical type
theories, obtained by implementation.
\item \textbf{Synthetic homotopy theory, effectively.}  The
$\pi_1(S^1)$ development realizes the Licata--Shulman encode--decode
method in a new implementation, and pushes further to identifications in
the universe ($\Omega S^1\equiv\Z$, $T^2\equiv S^1{\times}S^1$).  The
figure-eight computations exhibit \emph{effective} covering-space theory:
invariants evaluated, not just constructed.
\item \textbf{The classifying-space programme.}  $K(G,1)$ with its
$\op{Codes}$ family is the level-1 core of ``groups are pointed connected
1-types''.  Everything needed --- $\op{isGroupoidHSet}$, univalence
$\eta$, the h-level tower --- was built from first principles inside the
object theory.
\item \textbf{Coherence made computational.}  The pentagon/triangle
development, and especially the \emph{negative} finding about untruncated
base cases, gives a sharp computational articulation of what coherence
data is: exactly the cells that do not come for free from Kan operations
at abstract types.
\item \textbf{Metatheoretic data points.}  The substitution-instability
of strictly-cancelling $\Z$, the necessity of lazy boundary collapse,
and the junk-tolerance requirement are small theorems-by-implementation
about cubical evaluators that complement the paper literature.
\end{enumerate}

\section{Relevance to open problems}\label{sec:open}

We state plainly: \emph{this project has not solved, and does not claim
to solve, any of the open problems below.}  Its value is as a platform on
which well-defined fragments can be formulated and machine-checked, and
as a source of computational evidence.  With that stated:

\subsection{The homotopy hypothesis}
For Kan complexes the homotopy hypothesis is classical; the open
formulations concern Grothendieck $\infty$-groupoids
(Maltsiniotis, Ara, Henry) and, inside type theory, the impossibility to
date of even \emph{stating} the full hypothesis internally, which reduces
to the open problem of defining semi-simplicial types in HoTT.  The
project bears on this in three ways.
\begin{enumerate}[leftmargin=1.7em]
\item \emph{The level-$1$ fragment as theorems.}  ``Groups $\simeq$
pointed connected 1-types'' decomposes into machine-checkable pieces:
$\pi_1(S^1)\cong\Z$ (done), $K(G,1)$ with $\pi_1(K(G,1))\simeq G$
(constructed; final checks running), $S^1\simeq K(\Z,1)$ (constructed),
and --- new --- \textbf{the dimension-1 homotopy hypothesis in full}:
for every pointed \emph{connected} 1-type $(A,a)$,
\[ K(\Omega(A,a),1) \;\equiv\; A, \]
via the realization map (`em1rec` with the identity loop map, whose
coherence cell is the composition filler) shown to be an equivalence:
$\Omega(\mathrm{cong})$ agrees with `encode` by a two-step chain, is
therefore an equivalence on loops, extends to all endpoints by
connectedness of $K(G,1)$ (itself proved with no group laws), and yields
contractible fibers through a pair-path whose second component is a
corrected singular filler.  The whole chain is constructed and compiles;
the heavy conversion checks run in the native verifier.  Likewise
$\pi_1(S^1\vee S^1) = F_2$ (the free group on two letters, as reduced
words) is fully constructed: the covering space, both winding directions
--- the kernel computes transport along even the conjugated right-circle
generators definitionally --- and both round trips, the encode direction
by word induction through a no-cancellation lemma.  ``Groupoids
$\simeq$ 1-types'' remains on the roadmap.
\item \emph{Two-level architecture.}  The known obstruction to
internalizing $\infty$-structures is infinite coherence; the current best
strategy is two-level type theory (Annenkov--Capriotti--Kraus--Sattler).
This project \emph{is} structurally a 2LTT: Lean is a strict metatheory
in which towers indexed by external $\N$ --- e.g.\ a van den Berg--Garner
style $\omega$-groupoid structure on each object-language type, as a Lean
function $n\mapsto$ (cell structure as kernel-checked terms) --- can be
defined and verified.  This is a concrete, feasible research direction
enabled by the codebase.
This direction is now \emph{begun}: the module `LibTower`
defines, as a Lean function of external $n$, the iterated loop space
$\Omega^n(A,a)$ together with its composition, untruncated associator,
Mac~Lane triangle and pentagon, the middle-four interchange, and (from
dimension 2) the Eckmann--Hilton braiding --- each
rung a one-line instantiation of the generic coherence cells, stated
uniformly in the ambient type and kernel-checked per level (verified for
$n \le 3$; the check is cheap at every level).  Every object-language
type thus demonstrably carries the first rungs of a van den
Berg--Garner-style weak $\omega$-groupoid structure.  The tower is
moreover \emph{functorial}: iterated `cong` gives loop maps
$\Omega^n f$ at every level with no conjugation (basepoints are
preserved definitionally), and the functor law
$\Omega^n(g \circ f) \equiv \Omega^n g \circ \Omega^n f$ holds by
`refl` at every level.
\item \emph{Computational evidence.}  Where the hypothesis has
computational content (winding numbers, transport along
identifications), the platform executes it.
\end{enumerate}

\subsection{Axiomatization of coherence}
Classically, coherence is settled for bicategories and tricategories and
open (as a comparison/axiomatization problem) in general.  The cubical
point of view reframes it: Kan operations generate all compositions and
their coherences \emph{uniformly}, and the question becomes which
coherence cells are definitional and which require construction.  The
project's contribution is a precise experimental map of that boundary:
groupoid-level coherence is a one-liner ($1$-types); untruncated
coherence requires bootstrapped cells whose exact list
($\op{assocRefl}$, unit collapses at abstract types, \dots) is now
identified.  Extending this map upward (pentagonators, \dots) on this
platform is mechanical-but-informative work that would feed any future
axiomatization: it distinguishes, with machine certainty, \emph{free}
coherence from \emph{axiomatic} coherence.

\subsection{Weak higher categories and the comparison problem}
The library realizes the low-dimensional fragment: (pre)categories,
the 2-category $\mathbf{Cat}$ with interchange, hom-set-truncated functor
categories, and the groupoid $\op{hSet}$.  The comparison problem for
definitions of weak $n$-category is untouched --- honestly, it is out of
reach of any current proof assistant for general $n$ --- but the platform
supports two relevant activities: (i)~formalizing particular definitions
at low $n$ and proving instances of comparisons; (ii)~using the
meta-level (Lean) to manage the coherence bookkeeping that blocks
internal statements, per the 2LTT remark above.

\subsection{What would constitute progress here}
Concrete, staged milestones this platform can realistically host:
$\pi_1(K(G,1))\simeq G$ official (running); $S^1\simeq K(\Z,1)$;
groupoids $\simeq$ 1-types; a meta-level $\omega$-groupoid tower on
object types (2LTT-style van den Berg--Garner); the untruncated pentagon
via bootstrapped cells; $\pi_1(S^1\vee S^1)\simeq F_2$; and, as a
long-range target, $\pi_2(S^2)\simeq\Z$ (Freudenthal-grade).

\section{Current frontier}
As of this writing: the $\pi_1(K(G,1))$ verification chain runs on the
native infrastructure ($\op{uaCompMul}$ verified; remaining checks in
progress); evaluator performance science continues (value sharing;
conversion caching); see \texttt{HANDOFF.md} in the repository for the
day-level state.  This document is maintained alongside the code and
updated at each substantive milestone.

\end{document}
